Comparing Cross Validation Methods for Penalized Cox Regression

### 2017-09-13 cross-validation.Rnw
Add deviance residuals

\section{Introduction}
Cox propotional hazard regression is one of the most commonly used statistical model for analyzing data with survival outcomes. Despite its popularity, the cox model has its own limitations due to the semiparametric nature of the model. Since the model only estimates relative risks, it is challenging to evaluate the model's predictive accuracy.

>>>>

\section{Introduction}
Cox propotional hazard regression is one of the most commonly used statistical model for analyzing data with survival outcomes. Despite its popularity, the cox model is also limited by its the semiparametric nature. Since the model only yields estimations of relative risks, it is challenging to evaluate the model's predictive accuracy.

### 2017-09-13 cross-validation.tex
Add deviance residuals

\par The most commonly used approach of conducting cross validation in Cox regression is proposed by Verweij elt. \citep{Verweij1993}. They proposed a cross-validated log likelihood as a measure for the model's predictive accuracy.  It is currently implemented in the glmnet package with the "grouped" option. Similarly, the glmnet package also includes an "ungrouped" option where the cross-validated likelihood is calculated only based on the observations from the test set. In this paper, we proposed two alternative ways to carry out cross validation. Instead of cross-validating over partial likelihood, we propose to cross validate over the linear predictors of the regression model and cross validate over the Martingale residual of the Cox model.

@@

   \begin{figure}[h]
    \centering
    \begin{minipage}[b]{0.5\textwidth}
      \centering
		  \includegraphics[height= 3.5cm ]{./figures/03.png}
      \caption{Cross Validated Linear Predictors}
     \end{minipage}
     \begin{minipage}[b]{0.5\textwidth}
      \centering
		  \includegraphics[height= 3.9cm ]{./figures/04.png}
      \caption{Cross Validated Martingale Residuals}
      \end{minipage}	
   \end{figure}

  \subsection{Cross Validated Martingale Residuals}

Since the biggest challenge of conducting cross validation for cox regression is due to the missingness of baseline hazard, we propose a second approach which involves estimating the actual baseline hazard $\hat{\Lambda}_{0}$. Based on the observations in the training set, a baseline hazard $\hat{\Lambda}_{0}$ can be estimated via Kalbfleisch and Prentice's method. Then for each observation j in the test set, the Martingale Residual can be calculated: \begin{equation}\hat{M_{j}} = d_{j} -\hat{\Lambda}_{0}(t_{j})e^{\hat{\eta}_{j}},\end{equation} where $d_{j}$ is the status of the jth observation and $t_{j}$ is the time component of the jth observation. After repeating this procedure for all K folds, we would obtain a set of matingale residuals for all observations. They are be pulled together by sum of squares as the cross validated error:
\begin{equation}CVE = \sum_{j}\hat{M}_{j}^2\end{equation}

@@

    $\lambda$s chosen by different cross-validation methods are compared in Figure 5. All values are from taking the mean of 200 replications. The black line represents the optimal $\lambda$. All four cross validation methods consistently select $\lambda$s larger than the optimal one. Since larger $\lambda$ leads larger penalties and fewer variables are selected, all four cross validation methods seem to be conservative in terms of selecting variables. Ungrouped cross-validated log likelihood and the cross-validated linear predictor seem to be more liberal than the other two methods. The cross-validated martingale residual is the most conservative one and its performance is quite off from the other three methods.

@@

In Figure 9, the Cross Validated Error is rescaled and plotted for all four methods. A $\lambda$ will be selected when CVE curve reaches its lower point. The blue line, which represents the linear predictor approach, it has more curvature near its lowest point. It is easier to pick out the minimum point for this blue curve. Hence this approach is better at picking out signals than the other three approaches. If we look at the curve for the cross-validated Martingale Residual, there's almost no signal there.

@@

\begin{itemize}
	\item Explanation for Martingale Residuals
\end{itemize}

>>>>

\par The most commonly used approach of conducting cross validation in Cox regression is proposed by Verweij elt. \citep{Verweij1993}. They proposed a cross-validated log likelihood as a measure for the model's predictive accuracy.  It is currently implemented in the glmnet package with the "grouped" option. Similarly, the glmnet package also includes an "ungrouped" option where the cross-validated likelihood is calculated only based on the observations from the test set. In this paper, we proposed two alternative ways to carry out cross validation. Instead of cross-validating over partial likelihood, we propose to cross validate over the linear predictors of the regression model and cross validate over the deviance residual \citep{Therneau1990}of the Cox model.

@@

   \begin{figure}[h]
    \centering
    \begin{minipage}[b]{0.5\textwidth}
      \centering
		  \includegraphics[height= 3.5cm ]{./figures/03.png}
      \caption{Cross Validated Linear Predictors}
     \end{minipage}
     \begin{minipage}[b]{0.5\textwidth}
      \centering
		  \includegraphics[height= 3.9cm ]{./figures/04.png}
      \caption{Cross Validated Deviance Residuals}
      \end{minipage}	
   \end{figure}

  \subsection{Cross Validated Deviance Residuals}

Since the biggest challenge of conducting cross validation for cox regression is due to the missingness of baseline hazard, we propose a second approach which involves estimating the actual baseline hazard $\hat{\Lambda}_{0}$. As is mentioned in Therneau et al.'s original paper on Martingale Residuals \citep{Therneau1990}, the baseline hazard can be used to compute the Martingale residuals and deviance residuals, which are commonly used for Cox regression model diagnostics.

\par The original idea of calcualting the Martingale and Deviance residuals were integrated into cross validation. Based on the observations in the training set, a baseline hazard $\hat{\Lambda}_{0}$ can be estimated via Kalbfleisch and Prentice's method \citep{Kalbfleisch2011}. Then for each observation j in the test set, the Martingale Residual can be calculated: \begin{equation}\hat{M_{j}} = d_{j} -\hat{\Lambda}_{0}(t_{j})e^{\hat{\eta}_{j}},\end{equation} where $d_{j}$ is the status of the jth observation and $t_{j}$ is the time component of the jth observation. The deviance residual can be calcualted from the Martingale Residuals: \begin{equation} d_{i} = sgn(\hat{M}_{i})(-2(\hat{M}_{i} + \delta_{i}log(\delta_{i} - \hat{M}_{i})))^{1/2}\end{equation}

After repeating this procedure for all K folds, we would obtain a set of deviance residuals for all observations. They are be pulled together by sum of squares as the cross validated error:
\begin{equation}CVE = \sum_{j}\hat{d}_{j}^2\end{equation}.

One advantage of using deviance residual is that it is symmetric while Martingale Residual is not. However, there does exist a numeric challenge when incorporating deviance residuals for cross validation. Since the baseline hazard function $\hat{\Lambda}_{0}$ is a step function from 0, the cumulative hazard at the first time point from 0 is always 0. Since deviance residual contains a term $log(\delta_{i} - \hat{M}_{i})$, where $\delta_{i} - \hat{M}_{i} = \hat{\Lambda}_{0}$, this term would be negative infinity for the first time point and lead to a numeric issue for whichever fold that contains the first time point. Our solution is to smooth out the step function at the beginning.

@@

    $\lambda$s chosen by different cross-validation methods are compared in Figure 5. All values are from taking the mean of 200 replications. The black line represents the optimal $\lambda$. All four cross validation methods consistently select $\lambda$s larger than the optimal one. Since larger $\lambda$ leads larger penalties and fewer variables are selected, all four cross validation methods seem to be conservative in terms of selecting variables. Ungrouped cross-validated log likelihood and the cross-validated linear predictor seem to be more liberal than the other two methods. The cross-validated Deviance residual is the most conservative one and its performance is quite off from the other three methods.

@@

In Figure 9, the Cross Validated Error is rescaled and plotted for all four methods. A $\lambda$ will be selected when CVE curve reaches its lower point. The blue line, which represents the linear predictor approach, it has more curvature near its lowest point. It is easier to pick out the minimum point for this blue curve. Hence this approach is better at picking out signals than the other three approaches. If we look at the curve for the cross-validated Deviance Residual, there's almost no signal there.

@@
